---
title: "PML_Assignment"
author: "Matteo Zucchetta"
date: "Saturday, May 23, 2015"
output: html_document
---

In this document data of Human Activity Recognition 
have been analysed with the aim of building a machine learning to predict
the 

After a little preprocessing of the dataset, in order to fix some of the problems
in the data, the most relevant information will be used as predictors for a ML algorithm, after data transformation (pca-based, performed on train data).



# Data Processing.
The dataset is loaded into R using the *read.csv* function.


```{r, cache = TRUE}
library(caret)

data_train <- read.csv("pml-training.csv")

data_test<- read.csv("pml-testing.csv")


```

The characteristics of the dataset are analyses with the *summary* function (not
shown, see code). 

```{r, echo = T, results = 'hide'}
summary(data_train)
```

Many variables have too many NAs to be included. First of all data are splitted
in a training and a testing set, using about 30% of the data for evaluation
purpose. Data containing too many NAs (not suitable for a impute pre-processing)
or with a near zero variance, are excluded from the analysis. Also factor colums
with a large number of empty observation are removed.

```{r, echo = T, results = 'as.is', cache = TRUE}
set.seed(3452)

inTrain <- createDataPartition(data_train$classe, p = 1/10, list = FALSE)
tr_df <-  data_train[inTrain, ]
test_df <-  data_train[-inTrain, ] 

whichNA <- ifelse(unlist(lapply(1:ncol(tr_df),function(x) length(which(is.na(tr_df[, x]))))) / nrow(tr_df) > 0.1, 0, 1)

tr_df <- tr_df[, which(whichNA == 1)]
test_df <- test_df[, which(whichNA == 1)]

whichEMPTY <- ifelse(unlist(lapply(1:ncol(tr_df),function(x) length(which(tr_df[, x] == "")))) / nrow(tr_df) > 0.1, 0, 1)

tr_df <- tr_df[, which(whichEMPTY == 1)]
test_df <- test_df[, which(whichEMPTY == 1)]

nzv <- nearZeroVar(tr_df[, c(8:59)], saveMetrics = T)
which(nzv$zeroVar) # non problem!
```


Data is checked for correlations:

```{r, echo = T, results = 'as.is', warning = F, message = F}

library(Hmisc)
clus<-varclus(as.matrix(tr_df[, -c(1:7, 60)]), similarity=c("pearson"), type=c("data.matrix"))
plot(clus,ylim=c(0,0.8));abline(h=1-0.7^2,col="red")#;abline(h=1-0.95^2,col="red",lty=3)
```



Many retained variables are highly correlated, so the are pre-processed with a
PCA, keeping the number of PCs needed to explain 95% of the variance.

```{r, echo = T, results = 'as.is', warning = F, message = T, cache = T}

preProc<- preProcess(tr_df[, -c(1:7, 60)], method = "pca", thresh = 0.95)
preProcVal <- predict(preProc, tr_df[, -c(1:7, 60)])
print(paste('n° PCs =', ncol(preProcVal)))
```


# Model Fitting

A Random Forest is trained, along with a single tree model, just for comparison.
The two algorithm are fitted after the exploration of the optimal values of the 
training parameters (*mtry* and *cp*), by means of a 5-fold Cross Validation
performed on the training set, to obtain an estimation of the accuracy of the
classification.

```{r, echo = T, results = 'as.is', warning = F, message = T, cache = T}

trl <- trainControl(classProbs=TRUE, repeats = 5, method = "cv", number = 5)

grid_rf <- expand.grid(.mtry = c(3:12)) # model selection parameter in this case mtry.
grid_rpart<- expand.grid(.cp = c(0.01, 0.05,  0.1, 0.2, 0.5, 0.8, 0.99)) # model selection parameter in this case cp

ntr <- 2000 # number of trees of the RF

# Both trainControl and expand.grid is provided by caret
set.seed(89)
mod_rf <- train(x = preProcVal, y = tr_df[, 60], method = "rf", tuneGrid=grid_rf, metric = c("Kappa"), trControl = trl, ntree = ntr)

mod_rf
varImpPlot(mod_rf$finalModel)

mod_rpart <- train(x = preProcVal, y = tr_df[, 60], method = "rpart", tuneGrid=grid_rpart, metric = c("Kappa"), trControl = trl)
mod_rpart


plot(mod_rpart$finalModel)
text(mod_rpart$finalModel)

results <- resamples(list(mod_rf, mod_rpart))

dotplot(results)

```

# Accuracy

The estimated accuracy (by mean of CV) is **`r round(mod_rpart$results$Accuracy[which.max(mod_rpart$results$Kappa)],2)`** for the single tree (rpart) and **`r round(mod_rf$results$Accuracy[which.max(mod_rf$results$Kappa)],2)`** for the Random Forest model. This 
is the expected out of bag accuracy. The data not used for training the model
can be used to obtain a more robust accuracy estimation.



```{r, echo = T, results = 'as.is', warning = F, message = T, cache = T}

test_set <- predict(preProc, newdata = test_df[, -c(1:7, 60)])
cm <- confusionMatrix(predict(mod_rf, newdata = test_set), test_df$classe)
print(cm)
```

The accuracy of classification, according to the testing set of data is: **`r round(cm$overall[1], 2)`**.
Finally, the prediction of the Random Forest model for the 20 observation left
out for the evaluation, are given in the next chunk.

```{r, echo = T, results = 'as.is', warning = F, message = T, cache = T}

out_test <- data_test[, which(whichNA == 1)]
out_test <- out_test[, which(whichEMPTY == 1)]

out_proc <- predict(preProc, newdata = out_test[, -c(1:7, 60)])

table(out_test$problem_id,predict(mod_rf, newdata = out_proc))

```